openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-12 paper/policar2020-opentsne.tex
Text edits.

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis (PCA),
multidimensional scaling, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received
much attention as it can address high volumes of data and reveal meaningful
clustering structure. Most of the recent reports on single-cell gene expression
data start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}The figure shows the progression of t-SNE
	embeddings following theoretical advances, which are all included in
	\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
	single-cell gene-expression profiles of mouse retinal cells from the
	study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly,
	the data in \textbf{(c)} contains 27,499 expression profiles from mouse
	retinal cells from Shekhar \textit{et
	al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
	embedding following the parameter choices from the original publication
	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
	shows poor preservation of the global organization of clusters,
	resulting from random initialization and an affinity model focused on
	preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE
	embedding, utilizing the latest theoretical advances and practical
	recommendations. The embedding is constructed using a multi-scale
	affinity model, preserving both short-range and long-range interactions
	between data points, and is initialized such that the global layout is
	meaningful. Unlike in \textbf{(a)}, the green and blue clusters
	representing different sub-types of amacrine and bipolar cells are now
	localized to the same regions of the space, indicating a higher level
	of similarity than to other cell types. \textbf{(c)} shows how existing
	t-SNE reference atlases can be used to embed new samples into existing
	embeddings. The positions of new data points correspond to cell types
	from the reference atlas.}
\end{figure*}

Despite its utility, t-SNE has often been criticized for poor scalability when
addressing large data sets, lack of global organization -- t-SNE focuses on
local clusters that are arbitrarily scattered in the low-dimensional space --
and the absence of theoretically-founded implementations to map new data into
existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. Linderman \textit{et al.}
sped-up the method through an interpolation-based approximation, reducing the
time complexity to be merely linear dependent on the number of
samples~\cite{linderman2019fast}. Kobak \& Berens proposed several techniques
to improve global positioning, including estimating similarities with a mixture
of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a
principled approach for embedding new samples into existing
visualizations~\cite{policar2019embedding}.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new samples into existing
	embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
	well-defined clusters and uncover global relationships between clusters,
	and
\item improved defaults for learning rate and number of
	iterations~\cite{belkina2019automated} which produce embeddings at a lower
	computational cost.
\end{enumerate}

@@

\opentsne\ is open-source and has become an integral part of the Python data
science ecosystem. The package has received over 600 Github stars and has been
downloaded hundreds of thousands of times, averaging in thousands of weekly
downloads. The source code is publicly available on Github at
\url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
\textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python
package managers.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence to suggest that
any data set can accurately be described using only two dimensions and that any
embedding onto a two-dimensional plane will inevitably lead to loss of
information. Therefore, it is beneficial to examine multiple embeddings, each
of which provides a different perspective into the topology and other
characteristics of the data.
\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
Warm colors correspond to excitatory neurons, while
cool colors represent inhibitory neurons. While different clusters of
excitatory and inhibitory neurons appear close to one another, all clusters
appear somewhat equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white-space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell sub-types.

@@

Exaggeration can highlight transitions between cell-states in
developmental studies. Standard t-SNE often produces embeddings with
clearly-defined, discrete clusters. We have shown that we can adjust the level
of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
discrete clusters are often undesired in developmental studies where we assume
cells follow a continuous path between cell states. To this end, other
embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2}
are often used, as they tend to better capture the continuity between cell
states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
in Fig.~\ref{fig:cao}.a, the developmental path between different time points
is difficult to observe due to a large number of sprawled out clusters in the
embedding. On the other hand, the developmental paths are easy to follow
between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
factor from $1$ to $4$.

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously annotated single-cell data set and embed data from a separate
experiment into the reference landscape. The reference data taken from Macosko
\textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from
mouse retinal cells. Embedding samples from a similar experiment focusing only
on bipolar retinal cells by Shekhar \textit{et
al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new
cells correctly map to the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases where cell identities are not known in
advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
embedding using annotated data from Hochgerner \textit{et
al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
cells from the mouse brain. The authors provide annotations for each cell, and
we can verify their accuracy by visualizing well-established gene markers for the
major cell types. We then embed cells from Harris \textit{et
al.}~\cite{harris2018classes}, where annotations are provided only for neuronal
cells. We quickly identify other non-neuronal cell types, including
oligodendrocytes and astrocytes. We use marker genes to validate that new cells
were correctly mapped into the reference landscape.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples
	into an existing, reference embedding. We construct a t-SNE embedding
	for the data from Hochgerner \textit{et
	al.}~\cite{hochgerner2018conserved} containing 24,185 developing,
	single cells from the mouse hippocampus. The data contains different
	types of neurons, supporting glia, and other vascular cells (upper
	left). Cell annotations are taken from the original publications. The
	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
	hippocampal cells collected by Harris \textit{et
	al.}~\cite{harris2018classes} using the previous embedding as a
	reference. In their study, Harris \textit{et al.} collected 6,971
	single-cells and focused on identifying different types of inhibitory
	neurons. However, almost half of the collected cells are not neurons
	and were left uncharacterized. Inspecting where these cells are
	embedded in the reference embedding (bottom left) reveals that in
	addition to inhibitory neurons, the data contains several supporting
	glial cells as well as a small population of endothelial cells. We can
	verify the accuracy of our approach by inspecting marker genes for the
	major cell types in the reference (top row) and embedded samples
	(bottom row).}
\end{figure*}

These two examples demonstrate how \opentsne\ can be used to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing already
existing cell atlases. This approach is extremely
general and is not limited to single-cell gene-expression data, but can be
applied to any tabular data sets from any domain of science.

@@

Versatility is another of  \opentsne s core design principles. Kobak \& Berens
recently introduced several recommendations and tricks to obtain better and
more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
of multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with larger data sets. \opentsne\ provides an
extremely flexible API, where these improvements can be incorporated in just a
few lines of code. Furthermore, \opentsne\ supports fully-custom affinity models,
enabling users to construct t-SNE embeddings on
non-tabular data, e.g. relational data, as the only affinity-model requirement
is some notion of similarity between data points. Finally, our comprehensive
callback system can be utilized to monitor and adapt different stages of the
optimization phase and has been used to construct visually appealing animations
of the t-SNE optimization process.

@@

One of t-SNE's common criticisms is poor scalability to large data sets
containing up to millions of data points~\cite{becht2019dimensionality}. These
criticisms are largely based on two factors. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was
exacerbated further by the fact that the most widely-used implementation of
t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime
performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is
perfectly capable of creating t-SNE embeddings of millions of data points in a
matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large
data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations,
including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
(v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3). We perform
benchmarks on two processors, one representing usage on a personal computer and
one on a high-performance computing platform. The Intel(R) Core i7 is commonly
found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are
found in high-performance computing machines. Benchmarks were run with the
original t-SNE parameters for 1000 iterations, as some implementations do not
allow their modification.

@@

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
\opentsne\ exhibit better scaling than their Barnes-Hut
counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
even though \textsf{FIt-SNE} and \opentsne\ both implement the same
approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
written in Python, which inevitably incurs some runtime overhead compared to
its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
than its Barnes-Hut competitors. Additionally, modern computer processors
generally contain multiple cores, allowing us to make use of multi-threading.
This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart when utilizing multiple cores. We speculate that this is could be because \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which is aggressively optimized on
Intel(R) Xeon(R) processors.

@@

Widespread adoption of novel techniques is almost universally accompanied by
easy-to-use and easy-to-install software. While the t-SNE implementation from
\textsf{scikit-learn} fits this requirement, its implementation is
unfortunately very slow, making it unsuitable for its application to large data
sets. Other C++ implementations such as \textsf{MulticoreTSNE} and
\textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide
precompiled binaries, requiring users to compile the software themselves. This
is problematic from a usability standpoint, as many users often lack the
knowledge to do so. This is especially problematic with Windows users, where
the operating system does not include a C++ compiler by default, and correctly
configuring the command line can be an arduous process.

\opentsne\ is a Python package that focuses on accessibility. We provide
precompiled binaries for all major Python versions on all major platforms,
making the installation process as seamless as possible. \opentsne\ can be
installed through the Python Package Index (PyPI) or through conda on the
conda-forge channel. \opentsne\ is designed to be familiar to Python users and
follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well
established in the Python data science ecosystem. \opentsne\ implements
multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
schemes, allowing it to be used on data sets containing millions of data
points. While the Python virtual machine introduces some performance overhead,
the runtime is comparable to that of its pure C++ counterpart --
\textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables
researchers to quickly experiment with different settings and easily
incorporate their own components into the software. We provide a full feature
list and comparison to other popular t-SNE implementations in
Table~\ref{tab:features}.

>>>>

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, in particular for methods that produce informative data visualizations. Popular approaches include 
principal component analysis (PCA),
multidimensional scaling, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. 
Among these, t-SNE lately received much attention as it can address high
volumes of data and reveal meaningful clustering structure. Increased
resolution and throughput of modern molecular assays has lead to frequent use
of t-SNE in diverse fields including, but not limited to, single-cell
transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}.

Reports on single-cell gene expression data, our running example,
often start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}The figure shows the progression of t-SNE
	embeddings following theoretical advances, which are all included in
	\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
	single-cell gene-expression profiles of mouse retinal cells from the
	study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly,
	the data in \textbf{(c)} contains 27,499 expression profiles from mouse
	retinal cells from Shekhar \textit{et
	al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
	embedding following the parameter choices from the original publication
	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
	shows no preservation of the global organization of clusters,
	resulting from random initialization and an affinity model focused on
	preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE
	embedding, utilizing the latest theoretical advances and practical
	recommendations. The embedding is constructed using a multi-scale
	affinity model, preserving both short-range and long-range interactions
	between data points, and is initialized such that the global layout is
	meaningful. Unlike in \textbf{(a)}, the green and blue clusters
	representing different sub-types of amacrine and bipolar cells are now
	localized to the same regions of the space, indicating a higher level
	of similarity than to other cell types. \textbf{(c)} shows how existing
	t-SNE reference atlases can be used to embed new samples into existing
	embeddings. The positions of new data points correspond to cell types
	from the reference atlas.}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE emphasizes
local clusters that are arbitrarily scattered in the low-dimensional space --
and the absence of theoretically-founded methods to map new data into
existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. Linderman \textit{et al.}
sped-up the method through an interpolation-based approximation, achieving linear complexity in the number of
samples~\cite{linderman2019fast}. Kobak \& Berens proposed several techniques
to improve global positioning, including estimating similarities with a mixture
of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a
principled approach for embedding new samples into existing
visualizations~\cite{policar2019embedding}.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new samples into existing
	embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
	well-defined clusters and uncover global relationships between clusters,
	and
\item improved defaults for learning rate and number of
	iterations~\cite{belkina2019automated} which produce embeddings at a low
	computational cost.
\end{enumerate}

@@

\opentsne\ is open-source and has become adopted by the wider community,
receeiving over 600 Github stars and averaging thousands of weekly downloads.
The source code is publicly available on Github at
\url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
\textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python
package managers.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence that
every data set can accurately be described by using only two dimensions, and any such embedding will therefore inevitably lead to loss of
information. It is thus beneficial to examine multiple embeddings, each
of which provides a different perspective into topology and other
data characteristics.

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four embeddings of the same data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.  First,
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell subtypes.

@@

Exaggeration can highlight transitions between cell-states in
developmental studies. Standard t-SNE often produces embeddings with
clearly defined, discrete clusters. We have shown that we can adjust the level
of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
discrete clusters are often undesired in developmental studies where cells are assumed to follow a continuous trajectory between cell states. To this end, other
embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity between cell
states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time
points is difficult to observe due to a large number of sprawled out clusters.
On the other hand, the development becomes easier to follow when the
exaggeration factor is increased from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data taken from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from
mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, 
we can correctly map to the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for
cell type classification in cases where cell identities are not known in
advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
embedding using labeled data from Hochgerner~\textit{et
al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
cells from the mouse brain. The authors assign a cell type to each cell, and
we can verify their accuracy by visualizing well-established gene markers for the
major cell types. We then embed cells from Harris~\textit{et
al.}~\cite{harris2018classes}, where labels are provided only for neuronal
cells. We quickly identify other non-neuronal cell types, including
oligodendrocytes and astrocytes. We use marker genes to validate that new cells
were correctly mapped into the reference landscape.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples
	into an existing, reference embedding. We construct a t-SNE embedding
	for the data from Hochgerner \textit{et
	al.}~\cite{hochgerner2018conserved} containing 24,185 developing,
	single cells from the mouse hippocampus. The data contains different
	types of neurons, supporting glia, and other vascular cells (upper
	left). Cell labels are taken from the original publications. The
	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
	hippocampal cells collected by Harris \textit{et
	al.}~\cite{harris2018classes} using the previous embedding as a
	reference. In their study, Harris \textit{et al.} collected 6,971
	single-cells and focused on identifying different types of inhibitory
	neurons. However, almost half of the collected cells are not neurons
	and were left uncharacterized. Inspecting where these cells are
	embedded in the reference embedding (bottom left) reveals that in
	addition to inhibitory neurons, the data contains several supporting
	glial cells as well as a small population of endothelial cells. We can
	verify the accuracy of our approach by inspecting marker genes for the
	major cell types in the reference (top row) and embedded samples
	(bottom row).}
\end{figure*}

These two examples demonstrate how \opentsne\ can be used to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing existing cell
atlases. This approach is general and not limited to single-cell gene
expression, but can be used with any tabular data set regardless of field.

@@

Versatility is another of  \opentsne s core design principles. Kobak \& Berens
recently introduced several recommendations and tricks to obtain better and
more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
of multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with larger data sets. \opentsne\ provides a flexible
API, where these improvements can be incorporated in just a few lines of code.
Furthermore, \opentsne\ supports custom affinity models, enabling users to
construct t-SNE embeddings on non-tabular data, e.g. relational data, as the
only affinity-model requirement is a notion of similarity between data points,
i.e. the distance matrix.  Finally, our comprehensive callback system can be
utilized to monitor and adapt different stages of the optimization phase and
has been used to construct visually appealing animations of the t-SNE
optimization process.

@@

One of t-SNE's common criticisms is limited scalability to large data sets
containing up to millions of data points~\cite{becht2019dimensionality}. These
criticisms are largely based on two factors. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was
exacerbated further by the fact that the most widely-used implementation of
t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes
performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is
perfectly capable of creating t-SNE embeddings of millions of data points in a
matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large
data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
(v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3). We perform
benchmarks on two processors, one representing usage on a personal computer and
one on a high-performance computing platform. The Intel(R) Core i7 is commonly
found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are
found in high-performance computing machines. Benchmarks were run for 1000 iteration with the original t-SNE parameters, as some implementations do not
allow their modification.

@@

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
\opentsne\ exhibit better scaling than their Barnes-Hut
counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
even though \textsf{FIt-SNE} and \opentsne\ both implement the same
approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
written in Python, which inevitably incurs some runtime overhead compared to
its C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
than its Barnes-Hut competitors. Additionally, modern computer processors
generally contain multiple cores, allowing us to make use of multi-threading.
This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart when utilizing multiple cores. We speculate that this is could be because \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which is aggressively optimized on
Intel(R) Xeon(R) processors.

@@

Widespread adoption of novel techniques is almost universally accompanied by
easy-to-use and easy-to-install software. While the t-SNE implementation from
\textsf{scikit-learn} fits this requirement, its implementation is
unfortunately very slow, making it unsuitable for its application to large data
sets. Other C++ implementations such as \textsf{MulticoreTSNE} and
\textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide
precompiled binaries, requiring users to compile the software themselves. This
is problematic from a usability standpoint, as many users often lack the
knowledge to do so. This often occurs with Windows operating system, which does
not include a C++ compiler, making the correct configuration cumbersome.

\opentsne\ is a Python package that focuses on accessibility. We provide
precompiled binaries for all major Python versions on all major platforms,
making the installation process as seamless as possible. \opentsne\ can be
installed through the Python Package Index (PyPI) or through conda on the
conda-forge channel. \opentsne\ is designed to be familiar to Python users and
follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well
established in the Python data science ecosystem. \opentsne\ implements
multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
schemes, allowing it to be used on data sets containing millions of data
points. While the Python virtual machine introduces some performance overhead,
the runtime is comparable to that of its C++ counterpart --
\textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables
researchers to quickly experiment with different settings and easily
incorporate their own components into the software. We provide a full feature
list and comparison to other popular t-SNE implementations in
Table~\ref{tab:features}.
